Echoes from the scattering of wavepackets on wormholes

### 2019-01-13 paper.tex
Add files via upload

\subsection{Scattering by a traversable wormhole}\label{subsec:wh} 
In this section, we study the dispersion of Gaussian wavepackets by a traversable wormhole, formed by the junction at $r_0>r_g$ of two Schwarzschild black hole solutions with equal mass. There is a discontinuity in $G_{\alpha\beta}$ such that at $r=r_0=20.0r_g$, any contracting congruence of geodesics in one side of the throat starts to expand in order to reach the other side, violating the weak energy condition. Such a configuration is known in the literature as the Morris-Thorne traversable wormhole \citep{doi:10.1119/1.15620}. 
The dynamics of the scattering problem is still given by the solutions of \eqref{eq:scalar_eq_mov} following the same decomposition in spherical modes as in \eqref{eq:ylm_decomp}. Hence, the waveform of the equation of motion for the spherical modes is given by
\begin{equation}
\left[-\partial_t^2+\partial_x^2-V^{\mathrm{wh}}_{\mathrm{scalar}}(x)\right]\mathcal{U}_{\ell m}(x,t) = 0,
\label{eq:wave_scalar_wh}
\end{equation}
and the effective potential $V^{\mathrm{wh}}_{\mathrm{scalar}}(x)$ yields
\begin{equation}
V^{\mathrm{wh}}_{\mathrm{scalar}}\left(x\right) = \left(1-\frac{r_g}{r\left(|x|\right)}\right)\left[\frac{\ell(\ell+1)}{r\left(|x|\right)^2}+\frac{r_g}{r\left(|x|\right)^3}\right].
\end{equation}   
\begin{figure}[t!]
\centering
\includegraphics[width=.3\textwidth]{figures/Geom_optics.pdf}
\caption{\label{fig:Geom_optics} Schematic reconstruction of the outgoing solution by a sequence of reflections and transmissions inside the potential cavity.}
\end{figure}
which is plotted in Fig.~\ref{fig:potential_wh} and coincides with the shape of the potential calculated in \citep{Cardoso:2016oxy}. Strictly speaking, we refer to $r\left(|x|\right)$ as the same inverse of the function mentioned in \eqref{eq:tortoise2} now evaluated at $|x|-r_0$. As we can see in Fig.~\ref{fig:potential_wh}, the new effective potential is merely a reflection of the potential barrier in Fig.~\ref{fig:Potential_BH} about the ordinate axis; thus, it is sensible to identify this system as a potential cavity built from two potential barriers with the reflection and transmission coefficients depicted in Fig.~\ref{fig:RT_scalar}. Furthermore, let us assume that an arbitrary incident pulse $\Phi_0$ propagates towards the cavity, it is, therefore, reasonable to approximate the spectrum of the asymptotic solution by a simple geometrical series of reflections and transmissions inside the cavity acting on the incident pulse, as shown in Fig.~\ref{fig:Geom_optics}. Following this scheme, the asymptotic solution can be approximated by
\begin{equation}
\Phi^{\mathrm{wh}}(\omega)|_{x\rightarrow+\infty}=\left[R+T\circ\sum_{i=0}R^{2i+1}\circ T\right]\circ\Phi_0(\omega).
\label{eq:reconst}
\end{equation}
It is not difficult to see that, in the hypothetic case of a perfectly reflective wall replacing the left potential barrier in Figs.~\ref{fig:potential_wh} and \ref{fig:Geom_optics}, the reconstructed pulse is instead given by 
\begin{equation}
\Phi^{\mathrm{f}}(\omega)|_{x\rightarrow+\infty}=\left[R+T\circ\sum_{i=0}R^i\circ T\right]\circ\Phi_0(\omega),
\label{eq:reconst_firewall}
\end{equation}
corresponding to the case of a firewall. We will not cover the features of the firewall solution in what remains of this paper.

@@

@@

The amplitude of the outgoing signals depends on the width of the incident Gaussian wavepacket, this is visible in Fig.~\ref{fig:asympt_two_sigmas} where we plot the asymptotic solutions for two different ingoing wavepackets: one with $\sigma=0.6495r_g$ in the left panel and a second one with $\sigma=5.196r_g$ in the right. After subtracting the outgoing solution for a black hole, the presence of a train of wavelets, widely known as echoes, is very clear. The signal is not featureless after the transient, as we can observe in the right panel of the same figure. Therefore, subtracting the black hole outgoing pulse (i.e., the case in which there is only one potential wall) turns to be a convenient way to clean the signal from back reflections due to the ``tails'' of the potential barrier, especially since in this case the echoes are four orders of magnitude smaller than the transient. In both panels, we plot the variable $\Delta\mathcal{U}_{20}|_{\mathrm{clean}}\equiv\mathcal{U}_{20}|_{\mathrm{original}}-\mathcal{U}^{\mathrm{bh}}_{20}$ in the right upper corner of the figures to represent the echoes and their net amplitude after removing backscattering effects.

@@

In Figs.~\ref{fig:odd_sigma_small} and \ref{fig:odd_sigma_large} we see the asymptotic behavior of two solutions with $\sigma=0.9185r_g$ and $\sigma=5.196r_g$, respectively. Observing that the amplitude of the echoes is not large in general, since it varies depending on the spectral content of the initial pulses, which are not the same in the case of initial Gaussian wavelets with different widths.
\par
Frequency dependent reflection and transmission coefficients can be calculated by studying a scattering problem with a single potential wall, as we noticed in subsection \ref{subsec:bh}, this is simply achieved by doing an algebraic inversion of the tortoise coordinate definition in \eqref{eq:tortoise2}: here the inverted function is evaluated in $x-r_g$ instead of $|x|-r_g$. The definition of the reflectivity and transmissivity coefficients is the same as in \eqref{eq:ref_and_trans}

@@

In the left panel of Fig.~\ref{fig:RT_reconst}, we show the reflection and transmission coefficients as functions of the frequency, noticing that the two curves intersect at $R^2=T^2=0.5$, as expected. The identity $R^2+T^2$ is approximately satisfied. As a next step of our analysis, we reconstruct the Fourier transform of the asymptotic pulse shown in Fig.~\ref{fig:odd_sigma_small}.  
\begin{figure*}[t!]
\centering
\subfigure{
\includegraphics[width=.45\textwidth]{figures/RT_omega_odd.pdf}} \,
\subfigure{
\includegraphics[width=.45\textwidth]{figures/Reconst_omega_odd.pdf}}
\caption{\label{fig:RT_reconst} Left panel: Reflection and transmission coefficients for $\sigma=0.9185r_g$ and the quadrupole $\ell,m=(2,0)$, the ``sweet spot'' in the frequency domain is located around the intersection at $R^2=T^2=0.5$. Right panel: Geometrical optics reconstruction is plotted in red, and it is compared with the Fourier transform of the asymptotic solution shown in Fig.~\ref{fig:odd_sigma_small}. Ignoring the low frequency peaks (introduced by the finite size of the simulation box), we notice that the reconstructed spectrum provides a good idea of the overall shape, but it does not reproduce the power in the frequency of the QNMs. Our results are not dramatically different for the solutions of $\Psi^{\mathrm{even}}$.} 
\end{figure*}
\par
We employed the definition of the geometrical optics approximation in \eqref{eq:reconst}, applied up to $i=0$, in the Fourier transform of the Gaussian incident wavepacket in order to obtain the reconstructed profile in the right panel of Fig.~\ref{fig:RT_reconst}. The signal reconstructed using the geometrical optics approximation provides a better representation of the total reflected pulse as the ingoing wavelet gets wider, and therefore, it has more power in lower frequencies.

@@

\begin{figure*}
\centering
\subfigure{
\includegraphics[width=.45\textwidth]{figures/Echo_sigma_odd_l_2.pdf}} \,
\subfigure{
\includegraphics[width=.45\textwidth]{figures/Echo_sigma2_odd_l_2.pdf}} 
\caption{\label{fig:echo_sigma_odd} Left panel: Amplitudes of the first four echoes of $\Psi^{\mathrm{odd}}_{20}$ as a function of $\sigma$ for $\ell=2$. Right panel: Relative amplitude of the first four echoes compared to the amplitude of the transient. The points represent the simulated double/single wall pairs used in our analysis, as in Fig.~\ref{fig:echo_sigma}. These results are not significantly different for $\Psi^{\mathrm{even}}_{20}$. At low widths, the transient decreases faster than the amplitudes of the echoes as $\sigma$ becomes smaller.} 
\end{figure*}
\section{Conclusions}\label{sec:conclusions}
In this paper, we studied the scattering of a test scalar and tensor wavepacket on a Morris-Thorne wormhole. Using a Gaussian wavelet as an incident initial condition, we showed the time dependent scattering solution of the quadrupole scalar and tensor spherical modes both inside and outside the potential cavities in Figs.~\ref{fig:sol_n_flux_wh} and \ref{fig:even_evolve}. Furthermore, after finding the transmission and reflection coefficients of the cavities in Fig.~\ref{fig:RT_scalar} and in the left panel of Fig.~\ref{fig:RT_reconst}, we used the geometrical optics approximation to reconstruct the shape of the Fourier transformed asymptotic solutions in Fig.~\ref{fig:rec_scalar} and in the right panel of Fig.~\ref{fig:RT_reconst}. Finding that the reconstruction is accurate up to the quasinormal mode peaks, which were also identified in the span of frequencies where these can be found.

We found that in general, the echoes do not have a large amplitude as we can see directly in the left panel of Fig.~\ref{fig:asympt_two_sigmas} and in Fig.~\ref{fig:odd_sigma_large}. Moreover, there is a thin band of preferred frequencies (and the widths of the corresponding ingoing Gaussian signals) that maximizes the amplitude of the echoes. The frequency band is centered around the ``sweet spot'' in which the coefficients of transmissivity and reflectivity overlap, and it is precisely where the QNMs peaks are squeezed in. We extended our analysis to find the range in which the incident width of the incident pulse maximizes the amplitude of the first four echoes, and how large is their amplitude compared to the transient. In Figs.~\ref{fig:echo_sigma} and \ref{fig:echo_sigma_odd}, we found that for low widths of the ingoing signal, as the widths become smaller, the transient decreases faster than the amplitude of the echoes. In the study of the gravitational wave scattering by a Morris-Thorne wormhole, we find small differences between the Regge-Wheeler and the Zerilli effective potentials, as depicted in the right panel of Fig.~\ref{fig:evenodd_potentials}. Such a difference will induce a small polarization of the signal emitted by the cavity, as shown in Fig.~\ref{fig:pol}.
\begin{figure}[t!]
\centering
\includegraphics[width=.45\textwidth]{figures/Polarimetry_echo_w_06495.pdf}
\caption{\label{fig:pol} Since phases are random, we compare the Hilbert envelopes of the two clean solutions for echoes. It is possible to see a very small excess in amplitude coming from the odd solution, which means that the outgoing signal might have a small net polarization even in the scenario of an incident test Gaussian wavepacket of tensor perturbations with no preferred polarization.}
\end{figure}

@@

\begin{acknowledgments}
We would like to thank Alex Zucca, Levon Pogosian and Michael Desrochers for their time and their valuable comments and discussions. This project was partly funded by the Discovery Grants program of the Natural Sciences and Engineering Research Council of Canada. JG is supported by the Billy Jones Graduate Scholarship, granted by the Physics Department at SFU. This research was enabled in part by support provided by WestGrid (\url{www.westgrid.ca}) and Compute Canada Calcul Canada (\url{www.computecanada.ca})
\end{acknowledgments}
\bibliography{bibliography.bib}

>>>>

\subsection{Scattering by a traversable wormhole}\label{subsec:wh} 
In this section, we study the dispersion of Gaussian wavepackets by a traversable wormhole, formed by the junction at $r_0>r_g$ of two Schwarzschild black hole solutions with equal mass. There is a discontinuity in $G_{\alpha\beta}$ such that at $r=r_0=20.0r_g$, any contracting congruence of geodesics in one side of the throat starts to expand in order to reach the other side, violating the weak energy condition. Such a configuration is known in the literature as the Morris-Thorne traversable wormhole \citep{doi:10.1119/1.15620}. 
The dynamics of the scattering problem is still given by the solutions of \eqref{eq:scalar_eq_mov} following the same decomposition in spherical modes as in \eqref{eq:ylm_decomp}. Hence, the waveform of the equation of motion for the spherical modes is given by
\begin{equation}
\left[-\partial_t^2+\partial_x^2-V^{\mathrm{wh}}_{\mathrm{scalar}}(x)\right]\mathcal{U}_{\ell m}(x,t) = 0,
\label{eq:wave_scalar_wh}
\end{equation}
and the effective potential $V^{\mathrm{wh}}_{\mathrm{scalar}}(x)$ yields
\begin{equation}
V^{\mathrm{wh}}_{\mathrm{scalar}}\left(x\right) = \left(1-\frac{r_g}{r\left(|x|\right)}\right)\left[\frac{\ell(\ell+1)}{r\left(|x|\right)^2}+\frac{r_g}{r\left(|x|\right)^3}\right].
\end{equation}   
\begin{figure}[t!]
\centering
\includegraphics[width=.3\textwidth]{figures/Geom_optics.pdf}
\caption{\label{fig:Geom_optics} Schematic reconstruction of the outgoing solution by a sequence of reflections and transmissions inside the potential cavity.}
\end{figure}

@@

which is plotted in Fig.~\ref{fig:potential_wh} and coincides with the shape of the potential calculated in \citep{Cardoso:2016oxy}. Strictly speaking, we refer to $r\left(|x|\right)$ as the same inverse of the function mentioned in \eqref{eq:tortoise2} now evaluated at $|x|-r_0$. As we can see in Fig.~\ref{fig:potential_wh}, the new effective potential is merely a reflection of the potential barrier in Fig.~\ref{fig:Potential_BH} about the ordinate axis; thus, it is sensible to identify this system as a potential cavity built from two potential barriers with the reflection and transmission coefficients depicted in Fig.~\ref{fig:RT_scalar}. Furthermore, let us assume that an arbitrary incident pulse $\Phi_0$ propagates towards the cavity, it is, therefore, reasonable to approximate the spectrum of the asymptotic solution by a simple geometrical series of reflections and transmissions inside the cavity acting on the incident pulse, as shown in Fig.~\ref{fig:Geom_optics}. Following this scheme, the asymptotic solution can be approximated by
\begin{equation}
\Phi^{\mathrm{wh}}(\omega)|_{x\rightarrow+\infty}=\left[R+T\circ\sum_{i=0}R^{2i+1}\circ T\right]\circ\Phi_0(\omega).
\label{eq:reconst}
\end{equation}
It is not difficult to see that, in the hypothetic case of a perfectly reflective wall replacing the left potential barrier in Figs.~\ref{fig:potential_wh} and \ref{fig:Geom_optics}, the reconstructed pulse is instead given by 
\begin{equation}
\Phi^{\mathrm{f}}(\omega)|_{x\rightarrow+\infty}=\left[R+T\circ\sum_{i=0}R^i\circ T\right]\circ\Phi_0(\omega),
\label{eq:reconst_firewall}
\end{equation}
corresponding to the case of a firewall. We will not cover the features of the firewall solution in what remains of this paper.

@@

The amplitude of the outgoing signals depends on the width of the incident Gaussian wavepacket, this is visible in Fig.~\ref{fig:asympt_two_sigmas} where we plot the asymptotic solutions for two different ingoing wavepackets: one with $\sigma=0.6495r_g$ in the left panel and a second one with $\sigma=5.196r_g$ in the right. After subtracting the outgoing solution for a black hole, the presence of a train of wavelets, widely known as echoes, is very clear. The signal is not featureless after the transient, as we can observe in the right panel of the same figure. Therefore, subtracting the black hole outgoing pulse (i.e., the case in which there is only one potential wall) turns to be a convenient way to clean the signal from back reflections due to the ``tails'' of the potential barrier, especially since in this case the echoes are four orders of magnitude smaller than the transient. In both panels, we plot the variable $\Delta\mathcal{U}_{20}|_{\mathrm{clean}}\equiv\mathcal{U}_{20}|_{\mathrm{original}}-\mathcal{U}^{\mathrm{bh}}_{20}$ in the right upper corner of the figures to represent the echoes and their net amplitude after removing backscattering effects. Notice that, in the upper corner of both figures, the amplitude of the echoes does not decrease exponentially with time due to the successive reflections inside the cavity.

@@

@@

In Figs.~\ref{fig:odd_sigma_small} and \ref{fig:odd_sigma_large} we see the asymptotic behavior of two solutions with $\sigma=0.9185r_g$ and $\sigma=5.196r_g$, respectively. Observing that the amplitude of the echoes is not large in general, since it varies depending on the spectral content of the initial pulses, which are not the same in the case of initial Gaussian wavelets with different widths. In analogy with the scalar case, we notice in the upper corner of both figures that the amplitude of the echoes does not decay exponentially in time.
\par
Frequency dependent reflection and transmission coefficients can be calculated by studying a scattering problem with a single potential wall, as we noticed in subsection \ref{subsec:bh}, this is simply achieved by doing an algebraic inversion of the tortoise coordinate definition in \eqref{eq:tortoise2}: here the inverted function is evaluated in $x-r_g$ instead of $|x|-r_g$. The definition of the reflectivity and transmissivity coefficients is the same as in \eqref{eq:ref_and_trans}

\begin{figure*}[t!]
\centering
\subfigure{
\includegraphics[width=.45\textwidth]{figures/RT_omega_odd.pdf}} \,
\subfigure{
\includegraphics[width=.45\textwidth]{figures/Reconst_omega_odd.pdf}}
\caption{\label{fig:RT_reconst} Left panel: Reflection and transmission coefficients for $\sigma=0.9185r_g$ and the quadrupole $\ell,m=(2,0)$, the ``sweet spot'' in the frequency domain is located around the intersection at $R^2=T^2=0.5$. Right panel: Geometrical optics reconstruction is plotted in red, and it is compared with the Fourier transform of the asymptotic solution shown in Fig.~\ref{fig:odd_sigma_small}. Ignoring the low frequency peaks (introduced by the finite size of the simulation box), we notice that the reconstructed spectrum provides a good idea of the overall shape, but it does not reproduce the power in the frequency of the QNMs. Our results are not dramatically different for the solutions of $\Psi^{\mathrm{even}}$.} 
\end{figure*}

In the left panel of Fig.~\ref{fig:RT_reconst}, we show the reflection and transmission coefficients as functions of the frequency, noticing that the two curves intersect at $R^2=T^2=0.5$, as expected. The identity $R^2+T^2$ is approximately satisfied. As a next step of our analysis, we reconstruct the Fourier transform of the asymptotic pulse shown in Fig.~\ref{fig:odd_sigma_small}.  We employed the definition of the geometrical optics approximation in \eqref{eq:reconst}, applied up to $i=0$, in the Fourier transform of the Gaussian incident wavepacket in order to obtain the reconstructed profile in the right panel of Fig.~\ref{fig:RT_reconst}. The signal reconstructed using the geometrical optics approximation provides a better representation of the total reflected pulse as the ingoing wavelet gets wider, and therefore, it has more power in lower frequencies.

@@

\begin{figure*}
\centering
\subfigure{
\includegraphics[width=.45\textwidth]{figures/Echo_sigma_odd_l_2.pdf}} \,
\subfigure{
\includegraphics[width=.45\textwidth]{figures/Echo_sigma2_odd_l_2.pdf}} 
\caption{\label{fig:echo_sigma_odd} Left panel: Amplitudes of the first four echoes of $\Psi^{\mathrm{odd}}_{20}$ as a function of $\sigma$ for $\ell=2$. Right panel: Relative amplitude of the first four echoes compared to the amplitude of the transient. The points represent the simulated double/single wall pairs used in our analysis, as in Fig.~\ref{fig:echo_sigma}. These results are not significantly different for $\Psi^{\mathrm{even}}_{20}$. At low widths, the transient decreases faster than the amplitudes of the echoes as $\sigma$ becomes smaller.} 
\end{figure*}

\begin{figure}[t!]
\centering
\includegraphics[width=.45\textwidth]{figures/Polarimetry_echo_w_06495.pdf}
\caption{\label{fig:pol} Since phases are random, we compare the Hilbert envelopes of the two clean solutions for echoes. It is possible to see a very small excess in amplitude coming from the odd solution, which means that the outgoing signal might have a small net polarization even in the scenario of an incident test Gaussian wavepacket of tensor perturbations with no preferred polarization.}
\end{figure}

\section{Conclusions}\label{sec:conclusions}
In this paper, we studied the scattering of a test scalar and tensor wavepacket on a Morris-Thorne wormhole. Using a Gaussian wavelet as an incident initial condition, we showed the time dependent scattering solution of the quadrupole scalar and tensor spherical modes both inside and outside the potential cavities in Figs.~\ref{fig:sol_n_flux_wh} and \ref{fig:even_evolve}. Furthermore, after finding the transmission and reflection coefficients of the cavities in Fig.~\ref{fig:RT_scalar} and in the left panel of Fig.~\ref{fig:RT_reconst}, we used the geometrical optics approximation to reconstruct the shape of the Fourier transformed asymptotic solutions in Fig.~\ref{fig:rec_scalar} and in the right panel of Fig.~\ref{fig:RT_reconst}. Finding that the reconstruction is accurate up to the quasinormal mode peaks, which were also identified in the span of frequencies where these can be found.

We found that in general, the echoes do not have a large amplitude as we can see directly in the left panel of Fig.~\ref{fig:asympt_two_sigmas} and in Fig.~\ref{fig:odd_sigma_large}, where we also observe that the amplitude of the echoes does not decay exponentially in time. In addition to this, we found that there is a thin band of preferred frequencies (and the widths of the corresponding ingoing Gaussian signals) where the amplitude of the echoes is maximized. Such a frequency band is centered around the ``sweet spot'' in which the coefficients of transmissivity and reflectivity overlap, and it is precisely where the QNMs peaks are squeezed in. We extended our analysis to find the range in which the incident width of the incident pulse maximizes the amplitude of the first four echoes, and how large is their amplitude compared to the transient. In Figs.~\ref{fig:echo_sigma} and \ref{fig:echo_sigma_odd}, we found that for low widths of the ingoing signal, as the widths become smaller, the transient decreases faster than the amplitude of the echoes. In the study of the gravitational wave scattering by a Morris-Thorne wormhole, we find small differences between the Regge-Wheeler and the Zerilli effective potentials, as depicted in the right panel of Fig.~\ref{fig:evenodd_potentials}. Such a difference will induce a small polarization of the signal emitted by the cavity, as shown in Fig.~\ref{fig:pol}.

@@

\begin{acknowledgments}
We would like to thank Alex Zucca, Levon Pogosian and Michael Desrochers for their time and their valuable comments and discussions. This project was partly funded by the Discovery Grants program of the Natural Sciences and Engineering Research Council of Canada. JG is supported by the Billy Jones Graduate Scholarship, granted by the Physics Department at SFU. This research was enabled in part by support provided by WestGrid (\url{www.westgrid.ca}) and Compute Canada Calcul Canada (\url{www.computecanada.ca}).
\end{acknowledgments}
\bibliography{bibliography.bib}
